\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-12 analysis/paper/paper.tex
review 2nd case study

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \DeclareUnicodeCharacter{2500}{-}

@@

\begin{CodeInput}
R> AIC(casebase_exponential)
R> AIC(casebase_time)
R> AIC(casebase_splines)
\end{CodeInput}
\end{CodeChunk}

@@

The next example shows how case-base sampling can also be used in the
context of a competing risk analysis. For illustrative purposes, we will
use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
main author's website, and it is also available as part of the
\pkg{casebase} package.

@@

\begin{CodeInput}
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

The data contains information on 177 patients who received a stem-cell
transplant for acute leukemia. The event of interest is relapse, but
other competing causes (e.g.~transplant-related death) were also
recorded Several covariates were also captured at baseline: sex, disease
type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
CR3), source of stem cells (bone marrow and peripheral blood, coded as
BM+PB, or only peripheral blood, coded as PB), and age. A summary of
these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
We note that the statistical summaries were generated differently for
different variable types: for continuous variables, we gave the range,
followed by the mean and standard deviation; for categorical variables,
we gave the counts for each category.

\begin{table}[ht]
\centering
\begin{tabular}{ccc}
  \hline
Variable & Description & Statistical summary \\ 
  \hline
Sex & Sex & M=Male (100) \\ 
   &  & F=Female (77) \\ 
  D & Disease & ALL (73) \\ 
   &  & AML (104) \\ 
  Phase & Phase & CR1 (47) \\ 
   &  & CR2 (45) \\ 
   &  & CR3 (12) \\ 
   &  & Relapse (73) \\ 
  Source & Type of transplant & BM+PB (21) \\ 
   &  & PB (156) \\ 
  Age & Age of patient (years) & 4--62 \\ 
   &  & 30.47 (13.04) \\ 
  Ftime & Failure time (months) & 0.13--131.77 \\ 
   &  & 20.28 (30.78) \\ 
  Status & Status indicator & 0=censored (46) \\ 
   &  & 1=relapse (56) \\ 
   &  & 2=competing event (75) \\ 
   \hline
\end{tabular}
\caption{Baseline characteristics of patients in the stem-cell transplant study.}
\label{tab:table1bmtcrr}
\end{table}

In order to try and visualize the incidence density of relapse, we can
look at the corresponding population-time plot. In Figure
\ref{fig:compPop1}, failure times associated with relapse are
highlighted on the plot using red points, while Figure
\ref{fig:compPop2} provides a similar population-time plot for competing
events.

Our main objective is to compute the absolute risk of relapse for a
given set of covariates. First, we fit a smooth hazard to the data; for
the sake of this example, we opted for a linear term for time:

@@

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
R>     data = bmtcrr, 
R>     ratio = 100, 
R>     time = "ftime")
\end{CodeInput}
\end{CodeChunk}

From the fit object, we can extract both the hazard ratios and their
corresponding confidence intervals:

As we can see, the only significant hazard ratio is the one associated
with the phase of the disease at transplant. More precisely, being in
relapse at transplant is associated with a hazard ratio of 3.92 when
compared to CR1.

Given our estimate of the hazard function, we can compute the absolute
risk curve for a fixed covariate profile. We performed this computation
for a 35 year old woman who received a stem-cell transplant from
peripheral blood at relapse. We compared the absolute risk curve for
such a woman with acute lymphoblastic leukemia with that for a similar
woman with acute myeloblastic leukemia. Figure \ref{fig:compAbsrisk}
shows these two curves as a function of time. This figure also shows the
Kaplan-Meier estimate fitted to the two disease groups (ignoring the
other covariates).

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R>                                      levels = levels(bmtcrr[,"Sex"])),
R>                       "D" = c("ALL", "AML"),
R>                       "Phase" = factor(c("Relapse", "Relapse"), 
R>                                        levels = levels(bmtcrr[,"Phase"])),
R>                       "Age" = c(35, 35),
R>                       "Source" = factor(c("PB", "PB"), 
R>                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R>                         method = "numerical", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

@@

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset \citeyearpar{knaus1995support}. The
SUPPORT dataset tracks death for individuals who are considered
seriously ill within a hospital. Imputation was conducted using the
default methods in the \pkg{mice} package in \proglang{R}, available on
CRAN. Before imputation, there was a total of 1000 individuals and 35
variables. After imputation, we have 474 observations where 67.5\% of
individuals in the study experience the event of interest. The SUPPORT
dataset will be used to demonstrate regularization within casebase,
while comparing their absolute risk predictions for a new individual. A
description of each variable can be found in Table \ref{tab:support1}
and a breakdown of each categorical variable in Table
\ref{tab:support2}. The data was downloaded from the main author's
website, and it is also available as part of the \pkg{casebase} package.

@@

This report was generated on 2019-12-11 15:59:22 using the following
computational environment and dependencies:

@@

#>  date     2019-12-11                  

@@

@@

@@

#>  glue          1.3.1      2019-03-12 [1]

@@

@@

@@

@@

@@

@@

@@

@@

@@

\begin{CodeOutput}
#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [bc801c3] 2019-12-11: Merge branch 'Jesse-Islam-master' into review
\end{CodeOutput}
\end{CodeChunk}

>>>>

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \DeclareUnicodeCharacter{2500}{-}

@@

\begin{CodeInput}
R> c("Exp." = AIC(casebase_exponential),
R>   "Weib." = AIC(casebase_time),
R>   "Splines" = AIC(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

In the next case study, we show how case-base sampling can also be used
in the context of a competing risk analysis. For illustrative purposes,
we will use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
main author's website, and it is also available as part of the
\pkg{casebase} package.

@@

\begin{CodeInput}
R> library(casebase)
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

The data contains information on 177 patients who received a stem-cell
transplant for acute leukemia. The event of interest is relapse, but
other competing causes (e.g.~transplant-related death) were also
recorded. Several covariates were captured at baseline: sex, disease
type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
CR3), source of stem cells (bone marrow and peripheral blood, coded as
BM+PB, or only peripheral blood, coded as PB), and age. A summary of
these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
We note that the statistical summaries were generated differently for
different variable types: for continuous variables, we gave the range,
followed by the mean and standard deviation; for categorical variables,
we gave the counts for each category.

\begin{CodeChunk}
\begin{table}[t]

\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
\centering
\begin{tabular}{lll}
\toprule
Variable & Description & Statistical summary\\
\midrule
\code{Sex} & Sex & M=Male (100)\\
 &  & F=Female (77)\\
\code{D} & Disease & ALL (73)\\
 &  & AML (104)\\
\code{Phase} & Phase & CR1 (47)\\
\addlinespace
 &  & CR2 (45)\\
 &  & CR3 (12)\\
 &  & Relapse (73)\\
\code{Source} & Type of transplant & BM+PB (21)\\
 &  & PB (156)\\
\addlinespace
\code{Age} & Age of patient (years) & 4--62\\
 &  & 30.47 (13.04)\\
\code{ftime} & Failure time (months) & 0.13--131.77\\
 &  & 20.28 (30.78)\\
\code{Status} & Status indicator & 0=censored (46)\\
\addlinespace
 &  & 1=relapse (56)\\
 &  & 2=competing event (75)\\
\bottomrule
\end{tabular}
\end{table}

First, we can look at a population-time plot to visualize the incidence
density of both relapse and the competing events. In Figure
\ref{fig:compPop}, failure times are highlighted on the plot using red
points for the event of interest (panel A) and blue dots for competing
events (panel B). In both panels, we see evidence of a non-constant
hazard function: the density of points is larger at the beginning of
follow-up than at the end.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{../figures/compPop-1}

\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study. A) The red points represent relapse events. B) The blue points represent the competing events.}\label{fig:compPop}
\end{figure}
\end{CodeChunk}

Our main objective is to compute the absolute risk of relapse for a
given set of covariates. First, we fit a smooth hazard to the data using
a linear term for time:

@@

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R+     Status ~ ftime + Sex + D + Phase + Source + Age, 
R+     data = bmtcrr, 
R+     ratio = 100, 
R+     time = "ftime")
\end{CodeInput}
\end{CodeChunk}

We will compare our hazard ratio estimates to that obtained from a Cox
regression. To do so, we need to treat the competing event as censoring.

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R+                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{table}[t]

\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
\centering
\begin{tabular}{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.77 & (0.45, 1.35) & 0.68 & (0.39, 1.19)\\
Disease & 0.58 & (0.32, 1.04) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.11 & (0.45, 2.76) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.58 & (0.41, 6.04) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 3.89 & (1.81, 8.35) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.48 & (0.49, 4.46) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

From the fit object, we can extract both the hazard ratios and their
corresponding confidence intervals. These quantities appear in Table
\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
identified by case-base sampling is the one associated with the phase of
the disease at transplant. More precisely, being in relapse at
transplant is associated with a hazard ratio of 3.89 when compared to
CR1.

Given the estimate of the hazard function obtained using case-base
sampling, we can compute the absolute risk curve for a fixed covariate
profile. We perform this computation for a 35 year old woman who
received a stem-cell transplant from peripheral blood at relapse. We
compared the absolute risk curve for such a woman with acute
lymphoblastic leukemia with that for a similar woman with acute
myeloblastic leukemia. We will estimate the curve from 0 to 60 months.

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R+                                      levels = levels(bmtcrr[,"Sex"])),
R+                       "D" = c("ALL", "AML"), # Both diseases
R+                       "Phase" = factor(c("Relapse", "Relapse"), 
R+                                        levels = levels(bmtcrr[,"Phase"])),
R+                       "Age" = c(35, 35),
R+                       "Source" = factor(c("PB", "PB"), 
R+                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R+                         method = "numerical", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

We will compare our estimates to that obtained from a corresponding
Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
semiparametric model for the cause-specific \emph{subdistribution},
i.e.~the function \(f_k(t)\) such that
\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
model allows to directly assess the effect of a covariate on the
subdistribution, as opposed to the hazard. For the computation, we will
use the \pkg{timereg} package \citep{timereg}.

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
R+                           const(Phase) + const(Source) + const(Age), 
R+                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{../figures/bmtcrr-risk-1}

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{CodeChunk}

Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
case-base sampling and the Fine-Gray model. As we can see, the two
approaches agree quite well for acute myeloblastic leukemia; however,
there seems to be a difference of about 5\% between the two curves for
acute lymphoblastic leukemia. This difference does not appear to be
significant: the curve from case-base sampling is contained within a
95\% confidence band around the Fine-Gray absolute risk curve (figure
not shown).

@@

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset \citeyearpar{knaus1995support}. The
SUPPORT dataset tracks death for individuals who are considered
seriously ill within a hospital. Imputation was conducted using
predictive mean matching and using the \pkg{mice} package in
\proglang{R}. Before imputation, there was a total of 1000 individuals
and 35 variables. After imputation, we have 474 observations where
67.5\% of individuals in the study experience the event of interest.
\textcolor{red}{How come we lose individuals during imputation?} The
SUPPORT dataset will be used to demonstrate regularization within
casebase, while comparing their absolute risk predictions for a new
individual. A description of each variable can be found in Table
\ref{tab:support1} and a breakdown of each categorical variable in Table
\ref{tab:support2}. The data was downloaded from the main author's
website, and it is also available as part of the \pkg{casebase} package.

@@

This report was generated on 2019-12-12 13:11:14 using the following
computational environment and dependencies:

@@

#>  date     2019-12-12                  

@@

#>  cowplot     * 1.0.0      2019-07-11 [1]

@@

#>  farver        2.0.1      2019-11-13 [1]

@@

#>  glue        * 1.3.1      2019-03-12 [1]

@@

#>  kableExtra  * 1.1.0      2019-03-16 [1]

@@

#>  labeling      0.3        2014-08-23 [1]

@@

#>  lava          1.6.5      2019-02-12 [1]

@@

#>  numDeriv      2016.8-1   2016-08-27 [1]

@@

#>  plyr          1.8.5      2019-12-10 [1]

@@

#>  reshape2      1.4.3      2017-12-11 [1]

@@

#>  timereg     * 1.9.4      2019-07-30 [1]

@@

#>  viridisLite   0.3.0      2018-02-01 [1]
#>  webshot       0.5.1      2018-09-28 [1]

@@

\begin{CodeOutput}
#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [c4f42a9] 2019-12-11: Merge branch 'review-glmnet' into review
\end{CodeOutput}
\end{CodeChunk}
